Calculating the Solvent Exposed Surface Area of Proteins with Intersecting Spheres

### 2020-06-06 Paper.tex

\begin{figure}[h!]
\caption{}
\centerline{\includegraphics[width=3in]{mask}}
\end{figure}
All points on $a_i$ within $d_i$ of the $P_{iq}$ are buried, and those bits in $m_{iq}$ are set to zero, with the rest set to 1. This string can be AND'ed with $b_i$ to produce a new $b_i$. This can be done for all spheres that intersect $a_i$ for each sphere in turn. The key insight that cuts down on the $nkl$ calculations is the fact that the $m_{iq}$ can be stored and retrieved later if the same string is ever needed. This is very likely to occur, because the $d_{iq}$ can be made independent of $r_i$ by removing the factor. Then $d_{iq}=\sqrt{2(1-\cos\theta_{iq})}$. Now we can calculate $m_{iq}$ for the range of $d_{iq}$ from 0-2.
\subsection{Richmond's Algorithm and the Gauss-Bonnet Theorem}
In a previous paper written by Richmond (1984), a different strategy is used to find the solvent exposed surface area of the atoms. He still represents the atoms as spheres with definite radius with centers at the atomic coordinates, but instead of using the Alpha Shapes, he made use of the Gauss-Bonnet theorem, a fundamental theorem in differential geometry which connects the topological properties of a manifold (specifically its Euler Characteristic) with its geometric properties (its curvature) (Rotskoff 2010, Schnell 2004). A standard statement of the theorem is 
\int_M KdA + \int_{\partial M} k_g ds = 2\pi\chi(M)
Here $\chi$ is the Euler Characteristic of $M$, and $K$ and $k_g$ are the Gaussian and geodesic curvatures of the manifold and the boundary respectively. For polygons, $\chi=V-E+F=1$.

@@

$\sigma$ is related to the area of the ball, which for a ball with radius $\varrho$ is $4\pi\varrho^2$. Because we only care about the exposed surface area, we need to multiply by $\sigma$, in order to only get the solvent-exposed surface area. We can also define a weighted area, which only differs from the unweighted area equation by the factor of some parameter $\alpha$, which represents some constant atomic solvation parameters that were predefined by Eisenberg and McLachlan (1986). So for just one ball $B_i$:
\begin{align*}
&A = 4\pi\sum \varrho_i^2\sigma_i
&E = 4\pi\sum\alpha_i\varrho_i^2\sigma_i
\end{align*}

@@

We can think of the sphere as representing how much of the solvent can interact with the atom. 
\begin{comment}
The state \textbf{z} of each ball defines a center and a radius. As stated before, this paper considers that the weighted (E) and unweighted (A) areas are transformations from the space containing z ($\R^{3n}$) to $\R$. Their derivatives at a particular \textbf{z} are linear maps. In other words, the areas are represented by the map f and their derivatives the derivative of f at \textbf{z}.
\end{comment}
Because the area derivatives are linear, they can be broken down into a sum of components. Both unweighted and weighted area derivatives are comprised of two motions, direction preserving and distance preserving, which can be derived separately and summed to get the SASA derivatives.

@@

Using this spherical rectangle shape, and the fact that we know the caps are rotating, the change in area is found to be the change in the distance between the center of the circles $\eta_j$ divided by the circumference of the set of balls in the larger sphere (which is $S_j$ in this case) times the area of the slice of $S_j$ between $p$ and $q$: 
\begin{align*}
dA_j = \frac{d\eta_j}{2\pi\varrho_j}4\pi\varrho_j\varrho_{ijk}\\
\end{align*}
$\frac{d\eta_j}{2\pi\varrho_j}$ acts like the change in the “width” of the spherical rectangle while $4\pi\varrho_j\varrho_{ijk}$ serves as the change in the “height” of the spherical rectangle. The area derivative of this larger sphere becomes:
\begin{equation*}
\frac{dA_j}{d\eta_j }= 2\varrho_{ijk}
\end{equation*}

@@

The weighted area derivative theorem, defined as $DE_z(\textbf{t}) = <\textbf{e},\textbf{t}>$ is defined similarly to the unweighted area derivative theorem, but multiplied by the solvent parameter $\alpha$:

\begin{align*}
&\textbf{e} = \sum_j \sigma_{ij}\cdot e_{ij}+ \sum_j \sum_k \beta_{ijk} \cdot e_{ijk} \\
&e_{ij} = \pi\left((\alpha_i\varrho_i +\alpha_j\varrho_j) - \frac{(\alpha_i\varrho_i  - \alpha_j \varrho_j)(\varrho_i^2 - \varrho_j^2)}{\zeta_{ij}^2}\right)\cdot u_{ij} \\
&e_{ijk} = 2\varrho_{ijk}\frac{\alpha_i\varrho_i - \alpha_j\varrho_j}{\zeta_{ij}}\cdot u_{ijk}
\end{align*}
Not surprisingly, $e_{ij}$ and $e_{ijk}$ depend on (respectively) $u_{ij}$ and $u_{ijk}$ as well.

Above, we described $\textbf{a}$ and $\textbf{e}$. Using these, we can then clearly state the area derivative equations. For the unweighted area derivative equation, the direction preserving component is the the fraction of the solvent-exposed surface times the derivative of the area: $\sigma_{ij}<a_{ij},t_i>$ and the distance preserving component is the same: $\beta_{ijk}<a_{ijk},t_i>$. We know that the unweighted area derivative is the sum of these two components:
\begin{equation}
DA_z(\textbf{t}) = \sum_{i=0}^{n-1}\sum_{j\neq i}\left(\sigma_{ij}<a_{ij},t_i> + \sum_{k\neq i,j} \beta_{ijk}<a_{ijk},t_i>\right)
\end{equation}
Just like before, the weighted area derivative is basically the same as the unweighted derivative: 
\begin{equation}
DE_z(\textbf{t}) = \sum_{i=0}^{n-1}\sum_{j\neq i}\left(\sigma_{ij}<e_{ij},t_i> + \sum_{k\neq i,j} \beta_{ijk}<e_{ijk},t_i>\right)
\end{equation}
These are the derivative equations that can then be tested on different molecules. It is important to note that there are cases where these equations will be discontinuous. We won't elaborate on these, but, for example, discontinuities can happen in a case where three spheres intersect at a circle. There are other cases of discontinuities, which are useful to know in order successfully to run this algorithm for a variety of molecules.

@@

@@

\end{thebibliography}
\end{document}

>>>>

\begin{figure}
\caption{Calculation of $d_{iq}$ and $\theta_{iq}$ for the purpose of calculating $m_{iq}$.}
\centerline{\includegraphics[width=3in]{mask}}
\end{figure}

All points on $a_i$ within $d_i$ of the $P_{iq}$ are buried, and those bits in $m_{iq}$ are set to zero, with the rest set to 1. This string can be AND'ed with $b_i$ to produce a new $b_i$. This can be done for all spheres that intersect $a_i$ for each sphere in turn. The key insight that cuts down on the $nkl$ calculations is the fact that the $m_{iq}$ can be stored and retrieved later if the same string is ever needed. This is very likely to occur, because the $d_{iq}$ can be made independent of $r_i$ by removing the factor $r_i$. Then $d_{iq}=\sqrt{2(1-\cos\theta_{iq})}$. Now we can calculate $m_{iq}$ for the range of $d_{iq}$ from 0-2. Once we store these masks, we will have covered all possible overlaps between spheres because we have gone through all angles $\theta_{iq}$.
This provides a method of calculating the SASA of the entire protein. We can simply look at every intersection with a given atom and lookup the appropriate mask (based on the angle $\theta_{iq}$. Then we scale it by the radius of the atom we are looking at. Then we can perform our AND operations and find the SASA for a given atom. Iterating over all atoms gives us our answer.
\subsection{Richmond's Algorithm and the Gauss-Bonnet Theorem}
In a previous paper written by Richmond (1984), a different strategy is used to find the solvent exposed surface area of the atoms. He still represents the atoms as spheres with definite radius with centers at the atomic coordinates, but instead of using voxels, he made use of the Gauss-Bonnet theorem, a fundamental theorem in differential geometry which connects the topological properties of a manifold (specifically its Euler Characteristic) with its geometric properties (its curvature). A standard statement of the theorem is

\begin{equation}
\int_M KdA + \int_{\partial M} k_g ds = 2\pi\chi(M)
\end{equation}
Here $\chi$ is the Euler Characteristic of $M$, and $K$ and $k_g$ are the Gaussian and geodesic curvatures of the manifold and the boundary respectively. For polygons, $\chi=V-E+F=1$.

@@

$\sigma$ is related to the area of the ball, which for a ball with radius $\varrho$ is $4\pi\varrho^2$. Because we only care about the exposed surface area, we need to multiply by $\sigma$, in order to only get the solvent-exposed surface area. We can also define a weighted area, which only differs from the unweighted area equation by the factor of some parameter $\alpha$, which represents some constant atomic solvation parameters that were predefined by Eisenberg and McLachlan (1986). So for just one ball $B_i$:

\begin{align*}
&A = 4\pi\sum \varrho_i^2\sigma_i
&E = 4\pi\sum\alpha_i\varrho_i^2\sigma_i
\end{align*}

@@

We can think of the sphere as representing how much of the solvent can interact with the atom. 
\begin{comment}
The state \textbf{z} of each ball defines a center and a radius. As stated before, this paper considers that the weighted (E) and unweighted (A) areas are transformations from the space containing z ($\R^{3n}$) to $\R$. Their derivatives at a particular \textbf{z} are linear maps. In other words, the areas are represented by the map f and their derivatives the derivative of f at \textbf{z}.
\end{comment}

Because the area derivatives are linear, they can be broken down into a sum of components. Both unweighted and weighted area derivatives are comprised of two motions, direction preserving and distance preserving, which can be derived separately and summed to get the SASA derivatives.

@@

Using this spherical rectangle shape, and the fact that we know the caps are rotating, the change in area is found to be the change in the distance between the center of the circles $\eta_j$ divided by the circumference of the set of balls in the larger sphere (which is $S_j$ in this case) times the area of the slice of $S_j$ between $p$ and $q$: 
\begin{align*}
dA_j = \frac{d\eta_j}{2\pi\varrho_j}4\pi\varrho_j\varrho_{ijk}\\
\end{align*}
$\frac{d\eta_j}{2\pi\varrho_j}$ acts like the change in the “width” of the spherical rectangle while $4\pi\varrho_j\varrho_{ijk}$ serves as the change in the “height” of the spherical rectangle. The area derivative of this larger sphere becomes:

\begin{equation*}
\frac{dA_j}{d\eta_j }= 2\varrho_{ijk}
\end{equation*}

@@

The weighted area derivative theorem, defined as $DE_z(\textbf{t}) = <\textbf{e},\textbf{t}>$ is defined similarly to the unweighted area derivative theorem, but multiplied by the solvent parameter $\alpha$:
\begin{align*}
&\textbf{e} = \sum_j \sigma_{ij}\cdot e_{ij}+ \sum_j \sum_k \beta_{ijk} \cdot e_{ijk} \\
&e_{ij} = \pi\left((\alpha_i\varrho_i +\alpha_j\varrho_j) - \frac{(\alpha_i\varrho_i  - \alpha_j \varrho_j)(\varrho_i^2 - \varrho_j^2)}{\zeta_{ij}^2}\right)\cdot u_{ij} \\
&e_{ijk} = 2\varrho_{ijk}\frac{\alpha_i\varrho_i - \alpha_j\varrho_j}{\zeta_{ij}}\cdot u_{ijk}
\end{align*}
Not surprisingly, $e_{ij}$ and $e_{ijk}$ depend on (respectively) $u_{ij}$ and $u_{ijk}$ as well.

Above, we described $\textbf{a}$ and $\textbf{e}$. Using these, we can then clearly state the area derivative equations. For the unweighted area derivative equation, the direction preserving component is the the fraction of the solvent-exposed surface times the derivative of the area: $\sigma_{ij}<a_{ij},t_i>$ and the distance preserving component is the same: $\beta_{ijk}<a_{ijk},t_i>$. We know that the unweighted area derivative is the sum of these two components:
\begin{equation}
DA_z(\textbf{t}) = \sum_{i=0}^{n-1}\sum_{j\neq i}\left(\sigma_{ij}<a_{ij},t_i> + \sum_{k\neq i,j} \beta_{ijk}<a_{ijk},t_i>\right)
\end{equation}
Just like before, the weighted area derivative is basically the same as the unweighted derivative: 
\begin{equation}
DE_z(\textbf{t}) = \sum_{i=0}^{n-1}\sum_{j\neq i}\left(\sigma_{ij}<e_{ij},t_i> + \sum_{k\neq i,j} \beta_{ijk}<e_{ijk},t_i>\right)
\end{equation}

These are the derivative equations that can then be tested on different molecules. It is important to note that there are cases where these equations will be discontinuous. We won't elaborate on these, but, for example, discontinuities can happen in a case where three spheres intersect at a circle. There are other cases of discontinuities, which are useful to know in order successfully to run this algorithm for a variety of molecules.

@@

\bibitem{Chaudhury}
S. Chaudhury, S. Lyskov \& J. J. Gray, "PyRosetta: a script-based interface for implementing molecular modeling algorithms using Rosetta," Bioinformatics, 26(5), 689-691 (2010).

@@

\end{thebibliography}

\listoffigures
\end{document}
